# Data Types

Using any programming language effectively requires the user to learn how that language handles and stores data. For our purposes, we will be most interested in getting comfortable with the major value, variable, and dataset formats offered by SAS and R.

This guide will make a special effort to describe common pitfalls peculiar to each language. These pitfalls can be unintuitive and have been known to cause new programmers plenty of headaches. With a bit of luck, we'll learn how to avoid them so we can concentrate on what matters most: conducting a sound analysis.

## Major Differences Between R and SAS

R and SAS are very different beasts.

R is what's known as a functional language, meaning it is built around the general principle that we create objects (e.g., a dataset) feed them to functions that perform particular operations on the input object and return some sort of output.

As a simple example, let's take the object `iris`, a dataset provided by default in every R installation, and feed it to a function that returns the first 6 rows:

```{r datatypes-iris-1}
head(datasets::iris)
```

We can also provide directions to the function in the form of arguments. For instance, let's have a look at the first 10 rows instead by setting the `n` argument accordingly:

```{r datatypes-iris-2}
head(datasets::iris, n = 10)
```

We'll talk about some other ways to work with various objects, but for now, it suffices to understand this general principle about functional programming.

The syntax for SAS is very different. An example should suffice to illustrate. Let's take a look at the `iris` dataset again, performing the same two operations as we did in R:

```{sas datatypes-iris-3-datastep, collectcode = TRUE}
proc import datafile="data/iris.csv"
  out=iris
  dbms=csv
  replace;
run;

proc print data=iris(obs=6); run;
proc print data=iris(obs=10); run;
```

SAS is built around **datasteps** and **procedures** (the `proc` statements we wrote in the code above). As we continue on, we will see how different these languages are, and we'll discuss some strengths and weaknesses of both.

## R: Value Types

## SAS: Value Types
