# Data Structures

Using any programming language effectively requires that the user learn how the language handles and stores data. For our purposes, we will be most interested in getting comfortable with the major value, variable, and dataset formats offered by R and, to a lesser extent, SAS.

This guide will make a special effort to describe common pitfalls peculiar to each language. These pitfalls can be unintuitive and have been known to cause new programmers plenty of headaches. With a bit of luck, we'll learn how to avoid them so we can concentrate on what matters most: conducting a sound analysis.

## Major Differences Between R and SAS

R and SAS are very different beasts.

R is what's known as a functional language, meaning it is built around the general principle that we create **objects** (e.g., a dataset) and "feed" them to **functions** that perform particular operations on the input object and return some sort of output.

<span class="marginnote definition">
**Object**
Any entity stored in an R environment. Can be a variety of types, including vector, matrix, data.frame, function, and many others.
</span>

<span class="marginnote definition">
**Function**
A particular type of object that takes other objects and arguments as inputs, performs a set of procedures, and returns an output.
</span>

As a simple example, let's take the object `iris`, a dataset provided by default in every R installation. Below, we feed it to a function, `head()` that inspects the `iris` dataset and returns the first 6 rows:

```{r datatypes-iris-1}
head(datasets::iris)
```

We can also provide directions to functions in the form of **arguments**. For instance, let's instead have a look at the first 10 rows by setting the `n` argument in the `head()` function:

<span class="marginnote definition">
**Argument**
A component of a function that allows the user to set a variety of options governing the function's behavior or output.
</span>

```{r datatypes-iris-2}
head(datasets::iris, n = 10)
```

We'll talk about some other ways to work with various objects, but for now, it suffices to understand this general principle about functional programming.

SAS's syntax is very different.

An example should suffice to illustrate. Let's take a look at the `iris` dataset again, performing the same two operations as we did in R:

```{sas datatypes-iris-3-datastep, collectcode = TRUE}
proc import datafile="data/iris.csv"
  out=iris
  dbms=csv
  replace;
run;

proc print data=iris(obs=6); run;
proc print data=iris(obs=10); run;
```

SAS is built around **datasteps** and **procedures** (the `proc` statements we wrote in the code above). As we continue on, we will see how different these languages are, and we'll discuss some strengths and weaknesses of both.

## R: Object and Value Classes

As statistical programmers, we will need to have a good handle on the major classes (types), including which are appropriate based on the goals of our analyses. Here we will focus on two sets of entities---values and objects---and their respective classes.

```{r value-class-table, echo = FALSE}
tibble::tribble(
  ~`Value Class`,
  ~Description,
  ~Examples,

  "Numeric",
  "A numeric value, which might be an integer or decimal",
  "400, 3.1415, 0.07",

  "Boolean",
  "True/False",
  "TRUE",

  "Character",
  "A value that contains a string of alphanumeric characters",
  "php2200, systolbp, hiv_pos",

  "Factor",
  "A variable usually used to denote a categorical variable. Groups can be ordered or unordered, but each is assigned an invisible numeric value.",
  "Ordered: High, Medium, Low; Unordered: Red, Blue, Green",

  "Date",
  "Tend to be somewhat difficult to work with and come in many formats. Later, we will focus specifically on manipulating, analyzing, and displaying dates.",
  "2005-09-01; 09012005; September 1, 2005",

  "Missing",
  "Missing values in R are usually stored as NA. In certain cases, it may be advisable to specify the class of the missing value.",
  "Standard: NA; Class-specific: _NA_character, _NA_numeric"
) %>%
  knitr::kable(caption = "Major data types (value classes) in R")
```

To conduct a statistical analysis, usually we will need to store values in various types of objects. These objects can be tailored toward a specific data type or may enable the user to store multiple data types.


```{r object-class-table, echo = FALSE}
tibble::tribble(
  ~`Object Class`,
  ~Description,

  "Vector",
  "A one-dimensional data structure capable of storing a series of values regardless of class.",

  "Matrix",
  "A data structure with rows and columns that can extend into _n_ dimensions. Matrices are able to store only one data type at a time. For example, all stored values in a numeric matrix would be numeric.",

  "List",
  "An _n_-dimensional structure capable of storing multiple data types and objects in complex ways.",

  "Data frame",
  "A special type of list capable of storing mixed data types in rectangular format. Data.frames are what we probably think of when imagining a dataset from a trial or observational study. We will have a look at some subtypes of data.frames such as data.tables and tibbles later in the guide."

) %>%
  knitr::kable(caption = "Major object classes in R")
```

## SAS: Value Formats

We will spend a little less time reviewing SAS value formats because the first of the mandatory introductory SAS training modules covers these in a fair amount of detail.

<dev-note>Select major datatypes and refer readers to the comprehensive datatype dictionary in the SAS 9.4 documentation</dev-note>
