# Statistical Analysis

```{r, include = FALSE}
pacman::p_load(magrittr, readxl, DT, summarytools, dplyr, tidyr)
```

## Setup

In this section, we'll provide some examples of basic statistical analyses you might be called upon to do at some point.

Note that the analyses you conduct as part of real research will almost certainly be more complicated than the code shown here. However, the hope is that you'll get enough of a feel to begin putting the pieces together as demanded by a given project.

### Data

If you'd like to try any of these examples on your own, we're using the NHEFS data set freely available at the [website](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/) for _Causal Inference: What If_ by Miguel A. Hernán and James M. Robins.

For examples based in R, we'll use the CSV file they provide, and the sas7bdat data set for our SAS examples.

The website itself contains extensive code examples in SAS, Stata, R, and Python---everything you could want!---so those who will be taking advanced epidemiologic or causal inference coursework should refer to those resources for guidance on how to implement relevant analyses. We'll cover a little bit here, but only briefly.

### NHEFS Codebook

```{r nhefs-codebook, echo = FALSE}
readxl::read_excel("data/NHEFS_Codebook.xls") %>%
  knitr::kable()
```

## Exploratory Data Analysis

First, let's read in our data and glance at it using a couple of different views.

```{r read-nhefs, message = F}
nhefs <- readr::read_csv("data/nhefs.csv")
head(nhefs)
tibble::glimpse(nhefs)
```

For now, let's focus on observations (rows) that don't have any missing data. To do this, let's get a summary of how much missingness we have across the `r format(nrow(nhefs), big.mark = ",")` observations.

```{r count-missing}
allvars <- names(nhefs)

df_miss <- data.frame(variable = allvars)

df_miss <- df_miss %>%
  dplyr::rowwise() %>%
  dplyr::mutate(n_miss = sum(is.na(nhefs[[variable]])))

print(df_miss, n = nrow(df_miss))
```

That looks a little hairy. Basically, all we did here was to create data.frame, save the NHEFS variable names to its first column, then look up each variable in the NHEFS data set and count the number of missing values.

However, we could have gotten this summary in a number of different ways.

```{r alt-missing-counts}
lapply(nhefs, function(x) table(is.na(x)))
sapply(nhefs, function(x) sum(is.na(x)))
```

Let's drop every row with a missing value for any variable

```{r omit-too-many}
nhefs2 <- na.omit(nhefs)
tibble::glimpse(nhefs2)
```

Oops, we have only `r nrow(nhefs2)` rows left in the data set! Perhaps we should reconsider, and pick several variables that will be of interest to us for our current purposes.

```{r subset-variables}
selectvars <- c(
  "age", "alcoholfreq", "asthma", "wt71",
  "smokeintensity", "sbp", "allergies"
)

nhefs3 <- na.omit(nhefs[, selectvars])
tibble::glimpse(nhefs3)
```

That's better. We've discarded `r nrow(nhefs) - nrow(nhefs3)` rows and focused on a subset of variables of interest.

Look back at the output we got from `glimpse()`. Are the variables in the formats we expect/need? I see one that we may need to update, depending on how we plan to use it. (Hint: Look back at the codebook for the variable summaries)

It seems that `alcoholfreq` is current being treated as a numeric variables, but really, it's a factor with 5 discrete levels. Furthermore, level 5 is Unknown, which we should probably treat as missing. We'll also use a factor format for `alcoholfreq`.

<span class="marginnote">We could leave `alcoholfreq` as a numeric variable and wrap it in the `factor()` function only when we needed to. In some cases, that approach might be preferable, in some cases not.</span>

```{r factor-check}
nhefs3$alcfreq <- as.factor(nhefs3$alcoholfreq)
table(nhefs3$alcfreq)
```

Let's replace those 5 missing values with NA's, too, and drop the corresponding rows from our data set. Let's also get rid of the original `alcoholfreq` variable and just hang on to the factor variable we made.

```{r, alcfreq}
# In words: where alcfreq in nhefs3 equals 5, assign NA
nhefs3$alcfreq[nhefs3$alcfreq == 5] <- NA

nhefs4 <- na.omit(nhefs3)
nhefs4$alcoholfreq <- NULL

tibble::glimpse(nhefs4)
```

Let's summarize our dataset:

```{r eda-stats}
# Continuous variables
contvars <- c("age", "wt71", "smokeintensity", "sbp")
summary(nhefs4[, contvars])

# Categorical variables
catvars <- c("asthma", "allergies", "alcfreq")
lapply(
  X = catvars,
  FUN = function(x) {
    summarytools::freq(nhefs4[x])
  })
```

<span class="marginnote">Based on this [tidy example](https://www.r-bloggers.com/quick-plot-of-all-variables/).</span>

A few plots may help us, too:

```{r eda-density, warning = FALSE, dev = "svg"}
library(ggplot2)
library(ggthemes)

nhefs4 %>%
  dplyr::select(all_of(contvars)) %>%
  tidyr::gather() %>%
  ggplot(aes(x = value)) +
  facet_wrap(~key, scales = "free") +
  geom_density() +
  theme_tufte(base_size = 16)
```

```{r eda-bars, warning = FALSE, dev = "svg"}
nhefs4 %>%
  dplyr::select(all_of(catvars)) %>%
  tidyr::gather() %>%
  ggplot(aes(x = value)) +
  facet_wrap(~key, nrow = 2, scales = "free") +
  geom_bar(stat = "count", width = 0.5, fill = "lightgray") +
  theme_tufte(base_size = 16)
```

Don't worry just now about the plot code. We'll be looking in some detail at the ggplot2 package later.


## Tabular Analysis

Soon you will be no stranger to two-by-two tables and contingency tables. Introductory statistics and epidemiologic methods courses will require you to analyze such tables, either to measure the strength of an exposure's effect on some outcome or to establish a statistical association between two variables.

Going back to our NHEFS data, let's say we're interested in a possible association between quitting smoking and weight loss. The `qsmk` variable is a binary indicator of whether or not an individual quit smoking between 1971 and 1982, while the `wt82_71` variable measures an individual's change in weight (kilograms) between 1971 and 1982.

Typically, we would not want to dichotomize a continuous variable like `wt82_71`, but for the sake of this example, let's begin by creating a binary indicator of whether a subject gained weight during this timeframe.

```{r tabanalysis-gain-wt82_71}

# check whether any values of wt82_71 are missing
any(is.na(nhefs$wt82_71))

# count the number of missing observations
# is.na() will return TRUE if an observation is missing and FALSE if it is not
# we can use sum() to add up the number of TRUE values because R will treat a
# logical variable as numeric for this purpose, where TRUE = 1 and FALSE = 0
sum(is.na(nhefs$wt82_71))

# create the new variable, called wt82_71_gain
nhefs$wt82_71_gain <- NA #initialize with NA
nhefs$wt82_71_gain[nhefs$wt82_71 > 0] <- 1
nhefs$wt82_71_gain[nhefs$wt82_71 <= 0] <- 0

# make sure we coded the new variable correctly
summary(nhefs$wt82_71[nhefs$wt82_71_gain == 1])
summary(nhefs$wt82_71[nhefs$wt82_71_gain == 0])
```

The summary statistics we generated above should comply with our expectations: those we marked as having gained weight all have positive values of `wt82_71`, while those we marked as not having gained weight all have negative values of `wt82_71`. 

Note, too, that we have 63 missing values in our new variable, just as we saw in the original outcome.

```{r tabanalysis-2x2}
with(nhefs, table(wt82_71_gain, qsmk, exclude = NULL))
```

The option `exclude = NULL` above includes missing values in the table output. We'll ignore them for simplicity.

After doing so, we're left with the following two-by-two table for analysis.

```{r tabanalysis-pretty-2x2}

nhefs %>%
  count(qsmk, wt82_71_gain) %>%
  na.omit() %>%
  pivot_wider(
    names_from = wt82_71_gain,
    values_from = n,
    names_pref = "wtgain"
  ) %>%
  knitr::kable()
```

A natural choice here would be to conduct a two-sample test for the equality of proportions, where we compare the proportion of subjects in the quit smoking group (`r with(nhefs %>% select(qsmk, wt82_71_gain) %>% na.omit(), round(mean(wt82_71_gain[qsmk == 1]), 3))`) who gained weight versus the proportion in the non-quit smoking group (`r with(nhefs %>% select(qsmk, wt82_71_gain) %>% na.omit(), round(mean(wt82_71_gain[qsmk == 0]), 3))`).

```{r tabanalysis-proptest}

ptest <- prop.test(c(299, 757), c(299 + 104, 757 + 406))
ptest

```

The output above, shows us the group proportions, the $\chi^2$ test statistic, and the corresponding *p*-value. We also get a 95% confidence interval for the difference in proportions.

To get a tidier output as a `tibble`, the `broom` package often proves useful:

```{r tabanalysis-broom-2x2}
broom::tidy(ptest)
```

We also good have given the table directly to the `prop.test()` function:

<span class="marginnote warning">
**Know Your Output**
Notice that the proportions and, as a result, the confidence interval given by the second `prop.test()` are different than the first, even though the $\chi^2$ statistic and the *p*-value are the same. The statistical test we used applied to the distribution of cell counts in the table. However, if we needed to report the difference in proportions and its confidence interval, the second method would have given us an incorrect estimate. Always be sure you understand functions outputs!
</span>

```{r tabanalysis-tabinput-proptest}
with(nhefs, prop.test(table(qsmk, wt82_71_gain)))
```

## Regression Models
