```{r eval = FALSE, include = FALSE}
source("setup.r")
```

# Getting Comfortable with Loops and Logic

## What's a Loop?

A loop is simply a type of algorithm, in this case, a procedure in which we direct R to iteratively check for a particular condition and, depending on the state of that condition, conduct a particular action or terminate the algorithm.

Imagine we are conducting a randomized controlled trial and we want to assign study participants randomly to either the treatment or control arm. We decide we'll flip a coin for each participant and assign them to the treatment group if the coin comes up heads and to the control group if it comes up tails. Assuming we have 100 participants to assign, the following text provides an informal description of a loop designed to carry out this task:

1. Pick individual
2. Flip coin
   - IF coin = heads, assign to TREATMENT
   - ELSE IF coin = tails, assign to CONTROL
3. IF any individuals not assigned, repeat from step 1, ELSE stop.

Here's the equivalent in code, where our coin flip is a random draw of 1 or 0:

```{r for-loop-intro}

# generate participant IDs
pid <- 1:100

# set up a vector to store treatment group assignment
rx <- vector()

# set random seed so we get same set of coin flips each time
set.seed(98765)

for (i in 1:100) {
  coinflip <- rbinom(1, 1, prob = 0.5)

  if (coinflip == 1) {
    rx[i] <- "treatment"
  } else {
    rx[i] <- "control"
  }
}

data.frame(pid, rx)
```

The code above demonstrates the use of a **for-loop**. We also could have used a **while-loop**, which would have started as `while (i <= 100)`.

We can also express our algorithm as a flowchart. Often, mapping algorithms visually can help us as we plan or design loops and statistical programs in general. Furthermore, they make communicating complex procedures much easier.

<iframe class="diagram" src="images/for-loop-diagram.html"></iframe>
<!-- ```{r for-loop-flowchart, echo = F}
htmltools::includeHTML("images/for-loop-diagram.html")
``` -->


Loops are important and powerful tools in statistical programming and, when combined with logical operations, provide you with a means to develop sophisticated, flexible code bases.

In this chapter, we'll focus on some simple implementations of loops, focusing separately on self-written loops like the one displayed above and **vectorized loops** in R. We'll also take a look at a couple of looping procedures in SAS.

## Why Loop?

Looping often arises as a practical solution to repeat some task on a series of objects.

Let's say we had a list of numbers and wanted to multiple each of them by 10. Pretending we don't know about R's [recycling capabilities](working-with-data.html#recycling), we might write a loop to conduct this task.

```{r for-loop-example1}

# generate a list of arbitrary numbers to operate upon
numlist <- seq(from = 10, to = 100, by = 10)

# first, we need to create an empty object in which to store our results
results <- vector()

# loop
for (i in 1:length(numlist)) {
  results[i] <- numlist[i] * 10
}

print(results)
```

<span class="marginnote">To gain a little intuition, run the following loops and observe the results: `for (i in 1:3) print(i)`, `for (i in 32:60) print(i)`.
</span>

The `i` referred to in the `for` statement is a variable R creates to keep track of the current iteration of the loop. For instance, if R is processing the first iteration, `i` is stored as the number we've initialized after the `in` statement---in this case, we initialized the loop at 1. When R is on the third iteration, `i` is stored in the background as 3.

Therefore, based on what we've already learned in prior chapters, we can break down the steps as follows:

1. Create a sequence of numbers increasing from 10 to 100 by 10, and store it to the object `numlist`. (What kind of object is `numlist`?)

2. Create the empty vector `results`.

3. For each iteration `i` in the range 1--10, do the following:

   1. Multiply the <em>i</em>th element in `numlist` by 10 and store the result in the <em>i</em>th element of `results`.

4. Print the vector of results.

### Curly Braces

Note that we don't always have to use curly braces in our loops. Because our loop above is quite short, we could have written it as:

```{r curly-braces, eval = FALSE}
for (i in 1:length(numlist)) results[i] <- numlist[i] * 10
```

However, the syntax above requires that all the code be written in a single line. Moving forward, we'll almost always use curly braces when using procedures that allow them. These include not only `for` loops but `if`/`else` statements and `function` definitions.


## Nested Loops, Conditional Logic, and Reinventing Wheels

Usually, we'll be using loops to conduct more complicated operations. Let's pretend R doesn't have all these nifty built-in tools to summarize our data or create new variables.

Imagine instead that we're tasked with developing summary statistics for both categorical and continuous variables. To do so, we'll have to do the following:

1. Loop through a list of variables
2. Check to see if the variable is categorical or continuous
3. Calculate summary statistics appropriate to the variable type

```{r nested-loop-setup, echo = FALSE, message = FALSE}
library(dplyr)
nhefs <- readr::read_csv("data/nhefs.csv")

nh <- nhefs %>%
  select(seqn, death, sex, age, race, smokeyrs, smokeintensity, marital)

nhl <- lapply(nh, function(x) x)
```

I've selected a handful of variables from the NHEFS dataset we've already worked with.

```{r peak-nhefs-df}
head(nh)
```

However, instead of storing the data in a data frame, I've stored it in a list  for the purposes of the current examples.

```{r nhefs-list}
head(nhl)
```
<span class="marginnote">[Chapter 6](statistical-analysis.html) might have some hints on a quick way to do so.</span>

None of these variables has any missing data, but if you don't trust me, check for yourself!

First, though, take a minute to look through the list object and make note of a few things. All we've done is taken each variable from the `nh` dataframe and stored it as a separate vector in the list called `nhl`.

No observations have been reordered, so if we took the 7th value from each vector stored in `nhl`, we should be able to reproduce the 7th row of `nh`.

```{r nhefs-compare-row}
list7 <- unlist(lapply(nhl, function(x) x[7]))
df7 <- unlist(nh[7, ])

cbind(list7, df7)
```

So far, so good.

Before we start writing up our code to generate some descriptive statistics for these variables, we should list the tasks we'll need to execute in order to do so.

You may have noticed in the header of `nh` that all of the variables are labeled as `dbl`, meaning they are numeric. However, we know that some of them should (or could) be treated as categorical.

Just to be safe, we can check to make sure the variables were saved as numeric in the list as well.

```{r list-var-classes}
cbind(unlist(lapply(nhl, class)))
```

With each variable class established, we might decide we need to complete the following list of tasks:

1. Inspect each variable and calculate the appropriate descriptive statistics. For the current example, we'll just say we want a frequency table for categorical variables and a five-number summary for continuous variables.

2. Label each variable according to how we treated it. As an arbitrary rule, we'll consider variables with more than 6 unique values to be categorical and all others to be continuous.

3. Store the results.

Considering this problem will allow us to introduce a couple of looping approaches we might consider. The first is a self-written loop similar to the one we already wrote. The next couple of approaches will used _vectorized_ loops, either with functions provided as part of the standard R installation or with some newer functions provided via the `purrr` package.

### Standard Loop

To begin, we're going to create a new list `nhl_sum` in which each of its elements is itself a list storing the original variable vector, the variable's class type, and the summary statistics we calculate.

```{r std-loop}
nhl_sum <- list()

for (i in 1:length(nhl)) {

  # reinitialize an empty temporary list at the beginning of each iteration
  templist <- list()

  # test the number of unique values in the ith element of nhl
  # if there are 6 or fewer unique values, treat as categorical
  # otherwise, treat as continuous
  if (length(unique(nhl[[i]])) <= 6) {

    templist[["type"]]  <- "categorical"
    templist[["stats"]] <- table(nhl[[i]])

  } else {

    templist[["type"]]  <- "continuous"
    templist[["stats"]] <- summary(nhl[[i]])

  }

  # save the list we created into the ith element of our final holding object
  nhl_sum[[i]] <- templist

  # name the list element
  names(nhl_sum)[i] <- names(nhl)[i]
}

print(nhl_sum)
```

<span class="marginnote">Note that we included `seqn`, which is the recipient ID. We could have written the loop to omit that variable, or we could have dropped its summary from `nhl_sum`.</span>

If we just want to peak at the `smokeintensity` statistics, for instance, we can access them directly from `nhl_sum`:

```{r}
nhl_sum$smokeintensity
```

### Vectorized Loops

We're not going to focus properly on vectorization, except to review an implementation or two from what's called the `apply` family of functions in R. In general, vectorized loops can operate quite a bit faster in certain circumstances, but they also provide a different approach to iteration.

<span class="marginnote">To go beyond the looping examples covered here, DataCamp has a detailed [tutorial](https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r) on writing loops and controlling their flow and output.</span>

For instance, let's achieve something similar to what we did in the prior section with a standard `for` loop. With the `for` loop, we had to tell R explicitly how many loops to do, and we had to tell it where to store our results in `nhl_sum` each time.

We might see some advantages to this approach: for one, we can read the code line by line and see the entire sequence of events. We also forced ourselves to be exact about what R was to do with each piece of information, which is good practice.

But R gives us a variety of flexible tools to work with, and the `apply` functions can provide some more elegant, less verbose,^["Verbose" essentially means we have to type a lot. You may also see functions use a setting called "verbose" to control how much output a function gives you, so the word is used as a bit of jargon in programming.] and easier-to-debug solutions.

Anyway, since this introduction to vectorized loops is getting a little verbose, let's skip to the good stuff: we can use the `lapply` function to get a similar result as we did with a `for` loop.

```{r lapply-loop}
nhl_sum2 <- lapply(nhl, function(x) {

  if (length(x) <= 6) {
    list(type = "categorical",
         stats = table(x))
  } else {
    list(type = "continuous",
         stats = summary(x))
  }
})

print(nhl_sum2)
```

Remember, what we wrote above is also a loop, but we can translate roughly into words what we told R to do:

1. Loop through each column in `nhl`, and apply a function to it.
1. The function is: if the column contains fewer than 6 unique values, return a list containing a string named `type` (set to "categorical") and a frequency table stored in `stats`.
1. Otherwise, return a list containing a string named `type` (set to "continuous") and the results of `summary()` stored in `stats`.

The "l" in `lapply` refers to "list", meaning the function will return to us a list. In each drawer will be stored the results of each iteration of the function---in other words, one of the lists we told R to make in the if-else statements.

We may look at the code above and concede it is more succinct than the standard `for` loop, but maybe it's a tad ugly because we specified an anonymous function^[**Jargon alert:** An anonymous function simply means that the function is not named or stored in our environment.] within the `lapply` function itself.

We could have done the following for the sake of readability:

```{r lapply-loop-2}
autosummary <- function(x) {

  if (length(x) <= 6) {
    list(type = "categorical",
         stats = table(x))
  } else {
    list(type = "continuous",
         stats = summary(x))
  }
}

nhl_sum3 <- lapply(nhl, autosummary)
print(nhl_sum3)
```

Other functions in the `apply` family include `apply` itself, `sapply`, `mapply`, and `tapply`. Each provides variations on a theme, handling different sorts of inputs and returning different object types.

<span class="marginnote">A full review of the `apply` family is beyond the scope of this introduction, and to be honest, while your humble author uses several of these functions all the time, he will never really remember how to use the base `apply` function properly.</span>

In fact, you've already seen some of these functions used in prior code examples that were focused on other aspects of R. Scroll through prior chapters and sections to see if you can spot them and figure out what they're doing.

### Nested Loops

Nesting in this context refers to placing loops within loops.^[Kind of like dreams within dreams in the movie _Inception_, except nested loops make sense!]

Loop logic takes a while to get used to, and you might find you don't need them all at often. Nonetheless, they're something we should make an attempt to familiarize outselves with.

Perhaps the following code will say in generated output what is difficult to say in words:

```{r nested-loops}

alpha <- LETTERS[1:4]
nums <- 1:10

for (i in 1:length(alpha)) {

  cat("<<", alpha[i], ">>\n")

  for (j in 1:length(nums)) {
    cat(paste0(alpha[i], nums[j]), " ")
  }

  cat("\n\n")
}

```

In words, we loop through two objects, `alpha` and `nums`, tracked by `i` and `j`, respectively.

We get to the first element of `alpha`, output it to the console using `cat()`, and then, we loop through `nums`, paste each object to the current `alpha` element, and output that string to the console. Once we've finished doing this for all the elements of `num`, we move to the next iteration and continue until we've exhausted the whole procedure for all elements in `alpha`.
