# Data Visualization


```{r dataviz_setup, include=F}
knitr::opts_chunk$set(dev = "svg",
                      fig.align = "center")
```

Look at your data!

Yes, that means look at your raw data by producing subsets, printing to the console, and all the other things we've been doing. But it also means that we should visualize our data.

The potential purposes for doing so are legion, but we can name a few for starters:

- To detect errors (e.g., miscoded observations)
- To detect missing data
- To detect your own coding errors (e.g., miscoded derivative variables)
- To see the variation of single variables and the covariation between pairs of variables
- To summarize a lot of quantitative information so that it can be digested by the reader quickly and efficiently
- To reveal relationships between exposure and outcome
- To diagnose statistical or computational procedures (e.g., regressions, algorithms)
- To make principled decisions about how to analyze your data

Before we continue, though, we should keep one thing in mind: **data visualization is not about making pretty pictures**. Saying so does not discount the importance of aesthetics in your visualizations. However, the aesthetics you employ should be chosen to make the message of the figure clear to your audience. In some cases, we might want to highlight a particular contrast between two variables. In others, we might want to visualize how a particular variable trends over time. In yet others, we may simply want to visualize features in our raw dataset.

Each of these objectives implies the usefulness of certain visualizations and the impropriety of others. In other words, we want the purpose of the visualization to dictate the means by which we display relevant data.

In this introduction, we will focus on a couple of broad concepts and code examples in order to get you up and running as quickly as possible. However, I highly recommend two (freely available) resources in particular:

1. [Data Visualization: A Practical Introduction](https://socviz.co/) by Kieran Healy

2. [Top 50 ggplot2 Visualizations - The Master List (with Full R Code)](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html) by Selva Prabhakaran

Keeping these bookmarked will probably cover about 90%^[Source: Thin air] of your data visualization needs.

## Base R examples

You may find that working with plotting in base R is more efficient when you want quick-and-dirty visualizations---for instance, while you're conducting exploratory data analysis.

### Boxplots

```{r base-r-bp, out.width="75%"}
# using built-in iris dataset
boxplot(iris$Petal.Width)
boxplot(Petal.Width ~ Species, data = iris)
```

### Density Plots

```{r base-r-density, out.width="75%"}
plot(density(iris$Petal.Width))
```

### Scatterplots

```{r base-r-scatter, out.width="75%"}
plot(iris$Petal.Width)

plot(Petal.Width ~ Petal.Length, data = iris)
```

### Scatterplot Matrix

```{r base-r-scattermatrix, out.width="100%", fig.asp=1}
pairs(iris)
```

### Histograms

```{r base-r-hist, out.width="75%"}
hist(iris$Petal.Width)
```

### Linear Regression Diagnostics

Base R and some packages provide built-in plotting methods for certain types of objects. Here, feeding an object of the class `lm` will return a series of linear regression diagnostics.

```{r base-linear-reg-diag, out.width="85%"}
## run a regression on built-in dataset ChickWeight
fit <- lm(weight ~ Diet, data = ChickWeight)
class(fit)

plot(fit)
```

If you run `plot(fit)` as above, you might find R irritating in that it will print these plots to your image device (image pane in RStudio) in succession, leaving you to flip between images. For a nicer output, you can output a grid of regression diagnostics.

```{r base-linear-reg-diag-grid, out.width="85%", fig.asp=1}
par(mfrow = c(2, 2))
plot(fit)
``` 

## ggplot2 Examples

Before we start, let's load a few of the packages we'll be using. 

```{r ggplot-loads}
library(ggplot2)  # plotting package
library(ggthemes) # a package that adds plot themes to ggplot2
library(magrittr) # use pipes "%>%"
```

Before, we get into the examples, I'm going to apply a universal theme for the subsequent plots:

```{r ggplot-themeset}
theme_set(theme_few(base_size = 16))
```

The `ggplot2` package in R is an excellent package for data visualization based on the _grammar of graphics_ concept.^[Leland Wilkinson. [The Grammar of Graphics](https://link.springer.com/book/10.1007/0-387-28695-0). Springer, 2005.] The basic idea is to divide plot features into conceptual layers defined by their function in the visualization. 

Here's a general schema for how this concept maps to plot components in `ggplot2`:

- **data:** the underlying data set and variables we wish to convert into plot aesthetics
- **plot layer:** geometric entities (_e.g._, shapes, lines), statistical transformations, positioning elements
- **scales:** scales applied to geoms to map additional data dimensions (_e.g._, a color gradient)
- **coordinates:** the coordinate system used to plot the data (_e.g._, Cartesian, polar)
- **facets:** specifications that control a grid of plots, showing the same plot within multiple discrete groupings

Note that I've assigned names to each element for the sake of introduction, but the descriptions themselves follow closely from Hadley Wickham's (`ggplot2` creator) own in the paper referenced here.^[Hadley Wickham (2010) A Layered Grammar of Graphics, _Journal of Computational and Graphical Statistics_, 19:1, 3-28, DOI: [10.1198/jcgs.2009.07098](https://doi.org/10.1198/jcgs.2009.07098)]

In the following examples I will focus on longform syntax with the `ggplot` function. Learning the ins and outs of `ggplot2` will give you more flexibility and control over your images, particularly when you need to produce publication-quality graphics. However, `ggplot2` provides the `qplot` function for quick visualizations, akin to the base R graphics depicted above. One nice thing about `qplot` the its default setting to choose a geometric mapping automatically (hereafter, "geom"). Below, the function automatically selects a scatterplot.

```{r ggplot-qplot}
qplot(x = Petal.Width, y = Petal.Length, data = iris)
```

However, we can specify the geom explicitly. Here we get a smoothed line:

```{r ggplot-qplot-geomspec}
qplot(x = Petal.Width, y = Petal.Length, data = iris, geom = "smooth")
```

The [`ggplot2` website](https://ggplot2.tidyverse.org/reference) contains extensive documentation on all of the package's functions, including many helpful worked examples.

All the examples below use the in-built `iris` data set. Let's recall what it looks like:

```{r iris-preview}
head(iris, n = 10)
```

<span class="marginnote warning">
**Strange Symbols**
Note the difference between the `magrittr` pipe `%>%`, which we use to feed our data set to the `ggplot()` function and the `+` symbol we use to bind together the lines describing plot elements. This distinction is mandatory, as `ggplot2` will not allow you to use the `magrittr` pipe to link plot layers.
</span>

### Boxplots

```{r ggplot-boxplot, fig.asp = 1}
iris %>%
  ggplot(aes(x = Petal.Width)) +
  geom_boxplot()

iris %>%
  ggplot(aes(x = Species, y = Petal.Width)) +
  geom_boxplot()
```

### Density Plots

```{r ggplot-density, fig.asp = 1, fig.width = 5}
iris %>%
  ggplot(aes(x = Petal.Width)) +
  geom_density()
```

```{r ggplot-density-color, fig.asp = 0.75}
iris %>%
  ggplot(aes(x = Petal.Width, color = Species)) +
  geom_density()
```

### Scatterplots

If we want to see the relationship between a flower's petal length and its petal width, we can make a scatterplot.

```{r ggplot-scatterplot-marginal, fig.asp = 1, fig.width = 5}
iris %>%
  ggplot(aes(x = Petal.Width, y = Petal.Length)) +
  geom_point()
```

Perhaps this relationship varies by species. In that case we need to differentiate each point based on the `Species` variable. We have multiple options to do so. Two that make immediate sense would be to differentiate species by color or by shape.

```{r ggplot-scatterplot-grouped, fig.asp = 0.75, fig.width = 7}

iris %>%
  ggplot(
    aes(
      x = Petal.Width,
      y = Petal.Length,
      color = Species
    )) +
  geom_point()

iris %>%
  ggplot(
    aes(
      x = Petal.Width,
      y = Petal.Length,
      shape = Species
    )
  ) +
  geom_point(size = 2)
```

### Faceted Plots

Another option would have been to create a separate scatterplot for each species. In the plot below, each section of the grid is called a **facet**. We'll also save an initial (unfaceted) plot so we can play with the facet settings a bit.

```{r ggplot-scatter-facet, fig.asp = 0.5}
petal_facet <- iris %>%
  ggplot(aes(x = Petal.Width, y = Petal.Length)) +
  geom_point()

petal_facet +
  facet_wrap(vars(Species))

```

If we wanted a single column, we could have asked `facet_wrap()` for one.

```{r ggplot-scatter-facet-1col, fig.asp = 3, fig.width = 3}
petal_facet +
  facet_wrap(vars(Species), ncol = 1)
```

In practice, you will choose particular plot designs to illuminate the aspects of your data you want to highlight.


### Bar Plots

To demonstrate the difference between two functions, `geom_bar()` and `geom_col()`, we'll create a binary variable that encodes whether a given observation had a petal length less than 2 or at least 2.

```{r ggplot-bar}

library(dplyr)

### make discrete categories for Petal.Length
### only to demonstrate geom_bar and geom_col
iris_new <- iris %>%
  mutate(
    PL.group = case_when(
      Petal.Length < 2 ~ "PL1",
      Petal.Length >= 2 ~ "PL2",
      TRUE ~ NA_character_
    )
  )

head(iris_new)
```

The `geom_bar()` function will count up the number of members in each group. 

```{r ggplot-bar1, fig.asp = 1, fig.width = 5}

iris_new %>%
  ggplot(aes(x = PL.group)) +
  geom_bar()
```

Let's make that a little prettier, shall we?

```{r ggplot-bar2, fig.asp = 1, fig.width = 5}

iris_new %>%
  ggplot(aes(x = PL.group)) +
  geom_bar(
    fill = "lightgray",
    color = "black",
    width = 0.5
  )
```

Often you'll find that you'll be summarizing your data to make certain types of plots. Let's say we had already counted the number of members per `PL.group` level ourselves, as so:

```{r ggplot-plgrp-counts}

iris_sum <- iris_new %>% count(PL.group)
iris_sum
```

To create the preceding plot, we would use `geom_col()` instead of `geom_bar()`:

```{r ggplot-geom-col, fig.asp = 1, fig.width = 5}

iris_sum %>%
  ggplot(aes(x = PL.group, y = n)) +
  geom_col(
    fill = "lightgray",
    color = "black",
    width = 0.5
  )
```

In most cases, you'll find more than one viable route to your chosen graphic.


### Scatterplot Matrix

Base R has a nice function to create scatterplot matrices. To do so with `ggplot2` we need to install another package called `GGally`. 

<span class="marginnote idea">
One wonderful think about `ggplot2` is that it boasts a universe of extensions written by R users to achieve specific plotting goals. While learning the ins and outs of `ggplot2` is fun (for me, anyway), you may find [extensions](https://exts.ggplot2.tidyverse.org) useful.
</span>

```{r ggplot-ggallymatrix, message = F, warning = F, fig.asp = 1}
GGally::ggpairs(iris)
```

This visualization is nice because it gives you a lot of information in a small space. The bottom row even contains embedded facets!
