---
title: "Computing Concepts for Public Health Researchers"
author: "Jason Gantenberg, <span class='deg'>PhD, MPH</span>"
date: "Last Compiled: `r Sys.Date()`"
site: bookdown::bookdown_site
citeproc: false
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: jrgant/StatsCompPH-BrownSPH
description: "Resources and tutorials for R and SAS analysis geared toward public health students."
---

```{r book_setup, include = FALSE}
pacman::p_load(tidyverse, data.table, SASMarkdown)

# Windows path
# saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
# sasopts <- "-nosplash -ls 75"

# Linux (Ubuntu) path
saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en"
sasopts <- "-ls 75"

knitr::opts_chunk$set(
  engine.path = list(sas = saspath, saslog = saspath),
  engine.opts = list(sas = sasopts, saslog = sasopts),
  comment = NA,
  dev = "svg"
)

```

# Learning Objectives

Statistical computing underlies much of the work we do as researchers. However, students tend to be left to their own devices to develop their programming abilities.

While teaching oneself is an indispensable part of the process, new researchers may be better equipped to do so provided a base from which to develop their computing skills. This book is intended to serve as such an introductory resource, targeted at incoming graduate students at the Brown University School of Public Health. 

While the content focuses on coding in R, some examples of SAS code appear to provide a frame of reference for students having completed their degree program's required SAS training modules (if applicable).


## Main objectives

Install R (and RStudio)

  - Include R packages that will be helpful to download as the tutorials progress

Understand some basic data types in both SAS and R

Understand how each language handles:

  - Variable names
  - Value formats
  - Missing data
  - Objects
  - Environments (as applicable)

Be able to conduct the following tasks in R:

  - Import/export data to and from common filetypes
      - Use common machine-readable filetypes
  - Manipulate variables
      - Assign variable names and value formats
      - Change variable names and value formats
      - Create new or derived variables
  - Statistical analysis
      - Generate contingency tables and conduct appropriate statistical tests on them
      - Implement basic regression analyses
          - Linear
          - Logistic
          - Relative risk (log-binomial, Poisson, quasi-Poisson)
          - Proportional hazards
          <!-- - Complex survey methods -->
          <!-- - Generalized estimating equations -->
  - Simulate data
  - Write simple loops
  - Write simple functions/macros

Produce publication-quality plots

  - Review image formats (JPG, PNG, EPS, WMF, EMF, PDF)
  - Know where to find information on principles of data visualization

Consider ways to make your analyses reproducible

  - RMarkdown
  - Pseudorandom number seeds

Resource repository

  - Excellent, mostly free resources on programming and data visualization

## Secondary objectives

Get a feel for how to conduct more advanced analysis

  - Propensity scoring
  - Inverse probability weighting
  - Parameteric *g*-formula
  - Multiple imputation (don't have experience with this myself)
  - Propensity score matching (don't have experience with this myself)

Get comfortable with implementing multiple strategies for the same task

  - R: `base` vs. `data.table` vs. `tidyverse`

Understand the benefits of competing strategies in prior bulletpoint

  - Working with large datasets
      - Processing speed
      - Memory management

  - Code readability vs. parsimony
      - Documentation and reproducibility
      - Literate programming
      - Macros and functions
      - Code maintenance
      - Don't Repeat Yourself (DRY Principle)


## Acknowledgements

<p style="font-weight:normal;">I'd like to thank Prof. Alison Field for the opportunity to develop this resource and Brown University School of Public Health for providing the resources with which to do so. In addition, I'd like to thank the following individuals, who provided valuable feedback during development: Adam Sullivan, Rachel Yorlets, Christopher Barry, and Hongseok Kim.</p>


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown"), "packages.bib")
```
